Field-informed Reinforcement Learning for Learning Large-Scale Collective Tasks

### 2023-05-19 paper.tex
bib ok and sec 2

On the other hand, the programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/esocc/ViroliDB13,DBLP:journals/corr/ViroliADPB16,DBLP:journals/tocl/AudritoVDPB19},  
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.
%
Field calculus includes main operators for expressing \emph{spatio-temporal} computations that allow for 
\emph{i)} the progression of values over time, achieved by transforming a field computed in a previous round into a new field; 
\emph{ii)} the exchange of data with neighbouring fields (typically referred to as \lstinline|nbr|), where received data is accessed by adjacent fields; and 
\emph{iii)} the conditional division of computation into distinct domains of collective computation. 
%
%We avoid presenting here the full details of these operators since they are not essential to the contribution. 
For further elaboration on the actual field calculus, please refer to the work of~\cite{DBLP:journals/jlap/ViroliBDACP19}.

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source agent and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. 
%
In ScaFi~\cite{casadei2022scafi} %(a scala implementation of aggregate computing)
 it is simply expressed as:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the field expansion. 
This pattern is called \emph{gradient-cast} (or \emph{G}) and is a fundamental building block for the coordination of swarm-like systems.
It is expressed as: 
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
%\ga{more description here}
where \texttt{source} is a Boolean field indicating the presence of a source agent, 
 \texttt{value} is the value to be cast, and 
 \texttt{acc} is the accumulation function that is applied to the value during the cast.

%\lukas{Lukas to review this section again.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ a graph  
where $E$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $f_v$.
For the sake of simplicity, 
 we thereafter describe $G_f$ as a graph that contains the feature set $f_v$ for each node $v \in V$.
 Note that, when we refer to $G_f$ and $G_o$ we are referring to the same graph $G$ but with different node features.
 Also, to access the feature set $f_v$ of a node $v \in V$ we use the notation $f_v$ or $G_f[v]$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $f_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. 
 The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
%
In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $\mathcal{N}_G(v)$,
 and then combining it with the node's current embedding $h_v$.
%
The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. 
 With each traversal of an edge $e \in E$ from $v$, the number of layers increased
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}_G(v)\right\}\right)   
\end{equation}

\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}
where $h_{v}^{k}$ is the embedding of node $v$ within the $k$-th layer, 
 $\mathcal{N}_G(v)$ is the set of neighbours of node $v$ computed from $E$.
%
$h_v^0$ is the initial embedding of node $v$ and it is usually set to the node's feature vector $f_v$.
%
The differential part comes into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
 and it could be a simple sum, max or sum of products. 
 More complex aggregation functions are available~\cite{pellegrini2020learning}.
%
Thereafter, we express the application of a \ac{GNN} to a graph $G_f$ as:
\begin{equation}
GNN(G_f) = \{h_v^{(k)}: v \in V, k \in \mathbb{N}\}
\end{equation} 
This formulation allows \acp{GNN} to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.
%\lukas{do we need definitions of $o_v$, $h_v^*$ over time $t$?}

@@

 \Ac{RL} has gained a lot of interest recently, 
 thanks to its successful application in various scenarios, 
 ranging from video games (such as Alpha Go~\cite{Silver2016Go} and Atari~\cite{Atari2016DQN}) 
 to chatbots (like ChatGPT~\cite{ChatGPT2023}). 
% 
In \ac{RL}, an \emph{agent} (i.e., a smart entity capable of making decisions) 
 performs \emph{actions} in an \emph{environment} (i.e., everything outside the agent) according to a \emph{policy}, 
 to maximise long-term \emph{reward signal} (or minimise a \emph{regret}).

@@

 which renders the system non-stationary and difficult to manage. 

@@

 the problem can be modelled through the SwarMDP model~\cite{DBLP:conf/atal/SosicKZK17} -- 
 an extension of the \ac{DecPOMDP}~\cite{Decpomdp2000} model for swarm-like systems.

@@

  \item $\pi: \mathcal{O} \rightarrow \mathcal{A}$ is the policy function, which maps the observations to the actions. It could be deterministic or stochastic.

@@

In our specific case, the agents can only interact with 1-hop neighbourhoods and are not directly influenced by other agent observations. We can therefore restrict the observation model as follows:

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\begin{equation*}
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\qquad
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
\end{equation*}
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation perceived at the time $t$: $o^v_t \in O$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022} if we set $f_v = o^v_t$.
\subsection{Motivation}
%The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
% \emph{manual design}, where the application is conceived and described through a programming language, 
% and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesise distributed controllers. 
%
%Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
As mentioned above, our work falls between automatic and manual approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
%
Compared to earlier works on field-based coordination, 
 our approach builds on the concept of co-fields~\cite{DBLP:journals/pervasive/MameiZL04}, 
 where agents exploit the field as a stigmergic medium to receive system-wide information and apply reasoning to this data.
%  
We present a subsequent approach where agent intelligence is synthesised through \ac{MAARL} and the use of \ac{GNN} to learn a local representation from the neighbourhood.
%
By learning a smart policy directly in the environment, the agent becomes capable of adapting its behaviour to new situations.

>>>>

On the other hand, the programming model is ruled by the \emph{field calculus}~\cite{DBLP:conf/saso/ViroliBDP15,DBLP:journals/tocl/AudritoVDPB19},  
 a core language that allows the expression of collective behaviours through the composition of \emph{functions} operating on fields.
%
Field calculus includes main operators for expressing \emph{spatio-temporal} computations that allow for 
\emph{i)} the progression of values over time, achieved by transforming a field computed in a previous round into a new field; 
\emph{ii)} the exchange of data with neighbouring fields (typically referred to as \lstinline|nbr|), where received data is accessed by adjacent fields; and 
\emph{iii)} the conditional division of computation into distinct domains of collective computation. 
%
%We avoid presenting here the full details of these operators since they are not essential to the contribution. 
For further elaboration on the actual field calculus, please refer to the work of~\cite{DBLP:journals/jlap/ViroliBDACP19}.

On these minimal operators, 
 it is then possible to build \emph{self-organizing} coordination blocks.
 One of the founding blocks is the \emph{gradient} essential for 
 \emph{information} flows~\cite{DBLP:conf/saso/WolfH07}.
%
This operator generates a numerical field that represents the minimum distance from a source zone. 
%
In other words, it maps a Boolean field (where ``true'' indicates the presence of a source agent and ``false'' indicates its absence) to a distance field that indicates the proximity to the nearest source. 
%
In the ScaFi~\cite{casadei2022scafi} implementation of field calculus is defined by function:
\begin{lstlisting}
def gradient(source: Boolean): Double
\end{lstlisting}
Along the gradient, it is possible to cast information within the system, accumulating value during the field expansion. 
This pattern is called \emph{gradient-cast} (or \emph{G}):
\begin{lstlisting}
def G[V](source: Boolean, value: V, acc: V => V): V
\end{lstlisting}
%\ga{more description here}
where \texttt{source} is a Boolean field indicating the presence of a source agent, 
 \texttt{value} is the value to be cast, and 
 \texttt{acc} is the accumulation function that is applied to the value during the cast.
%
This is a fundamental building block for the coordination of swarm-like systems: in fact, the resulting system can be used to broadcast information, or route towards selected nodes.

%\lukas{Lukas to review this section again.}
\subsection{Graph Neural Networks}
%Over the years, several different neural networks have been proposed to solve specific tasks, ranging from 
%  simple \ac{MLP}~\cite{werbos1975beyond} to \ac{CNN}~\cite{oshea2015introduction}.
%  \lukas{more examples? }
%%
%Whereas \ac{CNN} is designed to be used with image and spatial-like data (e.g., audio), 
 \ac{GNN} is a novel neural network model used to process graph-structured data with deep learning approaches.
%
Let $G = (V, E)$ be a graph  
where $E\subseteq V\times V$ defines the neighbourhood relations for each participating node, 
and $V$ identifies the nodes present in the graph. %at that time.  \lukas{no time in the definition}
% 
Each node $v \in V$ is associated with an observation (or feature set) $f_v$.
For the sake of simplicity, 
 we thereafter describe $G_f$ as a graph that contains the feature set $f_v$ for each node $v \in V$.
 Note that, when we refer to $G_f$ and $G_o$ we are referring to the same graph $G$ but with different node features.
 Also, to access the feature set $f_v$ of a node $v \in V$ we use the notation $f_v$ or $G_f[v]$.
%\lukas{Over time this graph $G$ can change and therefore we define $G_t$ as the graph $G$ at a specific time $t$.} 
%
Given $f_v$, %a feature vector associated with each node $v \in V$, 
 the goal of a \ac{GNN} is to learn the node embedding $h_v$ for each node $v \in V$. 
 The node embedding $h_v$ describes the node in the network and summarises the geometric properties of the graph in this location, allowing for comparison of various nodes in the graph.
%
In modern \acp{GNN}, the node embedding $h_v$ is computed by aggregating information from the node's neighbours $\mathcal{N}_G(v)$,
 and then combining it with the node's current embedding $h_v$.
%
The graph is further partitioned into layers with the node $v$ at its bottom layer $0$. 
 With each traversal of an edge $e \in E$ from $v$, the number of layers increases.
Formally, a \ac{GNN} can be defined by two phases:
\begin{equation}
a_{v}^{(k)}=\operatorname{AGGREGATE}^{(k)}\left(\left\{h_{u}^{(k-1)}: u \in \mathcal{N}_G(v)\right\}\right)   
\end{equation}

\begin{equation}
h_{v}^{(k)}=\operatorname{COMBINE}^{(k)}\left(h_{v}^{(k-1)}, a_{v}^{(k)}\right)
\end{equation}
where $h_{v}^{k}$ is the embedding of node $v$ within the $k$-th layer, 
 $\mathcal{N}_G(v)$ is the set of neighbours of node $v$ computed from $E$.
%
$h_v^0$ is the initial embedding of node $v$ and it is usually set to the node's feature vector $f_v$.
%
The differential part comes into play in the $\operatorname{COMBINE}$ function, 
 which is usually a differentiable function such as a neural network layer.
%
$\operatorname{AGGREGATE} $ instead is a function that aggregates the information from the neighbours of a node $v$
 and it could be a simple sum, max or sum of products. 
 More complex aggregation functions are available~\cite{pellegrini2020learning}.
%
Thereafter, we express the application of a \ac{GNN} to a graph $G_f$ as:
\begin{equation}
\mathit{GNN}(G_f) = \{h_v^{(k)}: v \in V, k \in \mathbb{N}\}
\end{equation} 
This formulation allows \acp{GNN} to effectively process and extract features from graph-structured data by iteratively aggregating and transforming information from the node's neighbours.
%\lukas{do we need definitions of $o_v$, $h_v^*$ over time $t$?}

@@

 \Ac{RL} has gained a lot of interest recently, 
 thanks to its successful application in various scenarios, 
 ranging from video games (such as Alpha Go~\cite{Silver2016Go} and Atari~\cite{mnih2015playing}) 
 to chatbots (like ChatGPT~\cite{ChatGPT2023}). 
% 
In \ac{RL}, an \emph{agent} (i.e., a smart entity capable of making decisions) 
 performs \emph{actions} in an \emph{environment} (i.e., everything outside the agent) according to a \emph{policy}, 
 to maximise long-term \emph{reward signal} (or minimise a \emph{regret}).

@@

 which makes the system non-stationary and difficult to manage. 

@@

 the problem can be modelled through the SwarMDP model~\cite{DBLP:conf/atal/SosicKZK17}---an extension of the \ac{DecPOMDP}~\cite{Decpomdp2000} model for swarm-like systems.

@@

  \item $\pi: \mathcal{O} \rightarrow \mathcal{A}$ is the policy function, which maps the observations to the actions: it could be deterministic or stochastic.

@@

In our specific case, the agents can only interact with 1-hop neighbours and are not directly influenced by other agent observations. We can therefore restrict the observation model as follows:

@@

This model can be used then to express the evolution of the system in time.
Specifically, starting from a global state $\mathcal{S}^P_t$, the next state $\mathcal{S}^P_{t+1}$ is defined as:
\mathcal{A}^P_t = \pi(\xi(S^P_t))
\mathcal{S}^P_{t+1} = \mathcal{T}(\mathcal{S}^P_t, \mathcal{A}^P_t)
Given a time $t$, the system can be also represented as a graph $G_t = (V_t, E_t)$, 
  where $E_t$ is built from $\mathcal{N}$.
  Each node is then decorated with the local observation perceived at the time $t$: $o^v_t \in O$. 
  Therefore, this graph can be used for being processed by a \ac{GNN} as done in~\cite{DBLP:conf/corl/TolstayaGPP0R19,tolstaya2020learning,DBLP:conf/icra/GosrichMLPYR022} if we set $f_v = o^v_t$.
\subsection{Motivation}
%The definition of distributed controllers for intelligent systems can be divided into two major approaches: 
% \emph{manual design}, where the application is conceived and described through a programming language, 
% and \emph{automatic design}~\cite{brambilla2013swarm}, which employs machine learning techniques to synthesise distributed controllers. 
%Our work lies between these two approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
As mentioned above, our work falls in between automatic and manual approaches, specifically in the areas of field-based coordination and \ac{MAARL} with the use of \acp{GNN}.
Compared to earlier works on field-based coordination, 
 our approach builds on the concept of co-fields~\cite{DBLP:journals/pervasive/MameiZL04}, 
 where agents exploit the field as a stigmergic medium to receive system-wide information and apply reasoning to this data.
We present a subsequent approach where agent intelligence is synthesised through \ac{MAARL} and the use of \ac{GNN} to learn a local representation from the neighbourhood.
By learning a smart policy directly in the environment, the agent becomes capable of adapting its behaviour to new situations.
